Studying public transit API query logs\\ to get an indication of travel flows

### 2016-01-06 paper.tex
Acted on RV's comments

@@

Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
Several stages of data processing are required to arrive at this outcome.
As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
\rv{why? and is it just you, or also the reader?}
\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found further on: could you change the intervals int he visualizations to what matches \cref{fig:average}?} 
Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
\rv{Be a bit more precise here: they are valid traffic, but not helpful for prediction of human behavior.}
Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
Spelling mistakes, and the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.

@@

The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle. 
Each node is represented by an arc where its length is proportional to the total volume of incoming and outgoing flows. 
Chords or curved line segments are drawn to connect nodes. 
The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another. 
In our implementation, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
Interactive filtering is introduced to simplify the visualization and details are provided in pop-up dialogue when upon mouse-over.

@@

Results from the analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
Three observations in particular prompted further analysis of the data in greater detail. 
First, there has been a steady increase in the number of queries made between 2012 and 2015. 
\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
A dinstinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively. 
Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.

@@

There are several inconsistent patterns in the data that do not correspond with an existing understanding of flows in Belgium.
In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API or a high concentration people who utilise route planning applications on a regular basis.
\rv{Earlier, you said you filtered bots. Is this still the case?}
Accordingly, the reverse is identified in Liege where an unusually large connection originating from a suburb towards Liege city prompted speculation as well as discussions on how such outliers should be treated or what may have led to their existence.

>>>>

\todo{Maybe a sub header here?}

@@

Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
Several stages of data processing are required to arrive at this outcome.
As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
\rv{why? and is it just you, or also the reader?}
\todo{Alvin: no longer a problem since i changed the text to remove any mention of exact hours. PC: we will have to explain ourselves why we'd choose different hours than what we've found further on: could you change the intervals int he visualizations to what matches \cref{fig:average}?} 
While queries made by automated user agents like search engine bots and data harvesters can be considered as valid traffic, they may may over represent connections between certain stations. In this regard, such queries have been excluded to improve the accuracy of our analysis. 
\rv{Be a bit more precise here: they are valid traffic, but not helpful for prediction of human behavior.}
Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
Spelling mistakes, and the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.

@@

The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle. 
Each node is represented by an arc where its length is proportional to the total volume of incoming and outgoing flows. 
Chords or curved line segments are drawn to connect nodes. 
The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another.
In our implementation of the chord diagram, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
Interactive filtering is introduced to simplify the visualization and details are provided in pop-up dialogue when upon mouse-over.

@@

Results from the analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
Three observations in particular prompted further analysis of the data in greater detail. 
First, there has been a steady increase in the number of queries made between 2012 and 2015. 
\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
A distinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively. 
Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.

@@

There are several inconsistent patterns in the data that do not correspond with an existing understanding of flows in Belgium.
In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API that were not flagged as bots or a high concentration people who utilise route planning applications on a regular basis.
\rv{Earlier, you said you filtered bots. Is this still the case?}
Accordingly, the reverse is identified in Liege where an unusually large connection originating from a suburb towards Liege city prompted speculation as well as discussions on how such outliers should be treated or what may have led to their existence.
